Opportunities for improvement in adult trauma patients admitted to the intensive care unit: A registry-based study

### 2024-03-25 manuscript.Rmd
25/3

@@

This was a registry-based study using all patients between 2014 and 2023
from the Karolinska University Hospital in Solna who had been admitted
to the ICU and who were reviewed regarding the presence of OFI by the
mortality and morbidity conference. Bi-variable and multivariable
logistic regression was used to assess the following patient and process
factors and OFI among this group of patients: gender, age, if the
patient was intubated or not and how many days, Revised Trauma Score
(RTS), Injury Severity Score (ISS), on call, time from arrival at the
hospital until first computed tomography (CT), days in the ICU, ASA
preinjury and survival after 30 days.

@@

@@

Albaaj et al. recently analyzed opportunities for improvement (OFI) in
relation to different patient and process factors based on data from the
Swedish trauma registry and data from the Karolinska University Hospital
in Solna. The study indicated that patients with moderate to severe
trauma are at the highest odds of OFI [@4]. According to SweTrau 2022
41,1% of patients with New Injury Severity Score (NISS) \> 15 who were
at a university hospital were admitted to the ICU. The proportion of
severely injured admitted to the ICU has decreased in recent years,
however; patients with NISS \> 15 still constitute a large proportion of
the trauma patients treated in the ICU. A way to measure functional
level after hospital discharge is Glasgow Outcome Scale (GOS). According
to GOS, among patients with NISS \> 15 17% were deceased, 20% suffered
severe disability, 49% suffered moderate disability and 14% had good
recovery in the university hospitals of Sweden [@3].

@@

@@

For patients where hospital values for systolic blood pressure,
respiratory rate or GCS score were missing, prehospital values were used
instead. If the patient was intubated we automatically assumed that GCS
was equal to 3 and RR was equal to 0. For the RTS-variable, each of the
variables GCS, SBP and RR were then converted to a RTS-code between 0
and 4 based on what the values of GCS, SBP and RR were respectively. RTS
was then calculated using the formula for RTS based on the combination
of the RTS-codes assigned to each of the previously listed variables.

@@

The study outcome will be on the presence of OFI in adult patients
treated in the ICU based on patient and process factors.

@@

The quantitative variables were factors chosen from the trauma registry,
based on the locally used audit filters, standard epidemiological
factors and factors registered in the Swedish Trauma Registry. The
categorical factors were sex, age, if the patient was intubated or not
and how many days, Revised Trauma Score (RTS), Injury Severity Score
(ISS), on call, time from arrival at the hospital until first computed
tomography (CT), days in the ICU, ASA preinjury and survival after 30
days. On duty was defined as before 8.00 a.m., after 5 p.m., or during a
weekend - defined as Saturday or Sunday.

@@

There were 14,022 patients in the trauma registry between January 1 2013
and February 28 2023. 2679 of these patients were admitted to the ICU. 3
patients were then excluded as they were younger than 15 years and 6
patients were excluded because they were dead on arrival. 928 patients
were excluded because there was no registration on whether there was an
OFI or not. 293 patients were excluded because they had missing values
in one or more variables. This leaves us with 143 outcomes for OFI, and
1,306 outcomes for "No OFI".

@@

Sex, Age, Intubation, ISS, time to first CT, on duty and ASA preinjury
were not significantly associated with OFI in the unadjusted or the
adjusted analyses. Female sex, intubation \> 7 days, \> 7 days in the
ICU, ASA preinjury 1-2 and survival 30 days after hospitalisation could
not be analysed with logistic regression.

@@

### Key results

We found that higher RTS was associated with higher odds of OFI. RTS is
a combined value of systolic blood pressure, Glasgow Coma Scale and
respiratory rate with more weight towards GCS. An explanation for the
increased odds of OFI in patients with increased RTS could be that the
patients with low RTS values - with low GCS, RR and SBP also are the
once that are in a bad state upon arrival --- upon which emergency
procedures e.g. intubation have been carried out rapidly due to the
patients being visibly ill from the beginning. The patients who are
being missed and where certain procedures take longer time before being
performed are the patients who seem stable upon arrival, but who later
become worse --- the patients with higher RTS scores. In Albaaj's study
respiratory rate \>29, systolic blood pressure 50-75 mmHg and 76-89
mmHg, and Glasgow Coma Scale 9-12 were significantly related to
increased odds of OFI in the unadjusted analysis, but not in the
adjusted analysis among patients from the Swedish Trauma Registry [@12].
All of these variables equal medium to higher RTS values --- capturing
the group of patients who's vital parameters are negatively affected by
the trauma, but who are relatively stable in this group of patients.
This creates a sense of false security, leading to increased time until
necessary emergency procedures are performed.

#Se om det finns källa som pekar på vilken grupp av patienter som bör få
mer fokus på IVA?

The variable ≤ 7 days in the ICU was significantly associated with
decreasing odds of OFI in both the adjusted and unadjusted analyses.
There are two aspects to consider concerning this variable. One aspect
would be that less days spent in the ICU compared to more days spent in
the ICU indicates that patients have been in a better state and needed
less time of recovery before leaving the hospital. Furthermore, a longer
stay leads to more space for mistakes being made. The other aspect of
the variable would be that few days in the ICU could indicate the
patient passing away before any mistakes having been made due to the
patient being very ill from the beginning --- the deaths were not
preventable, but rather expected.

#Leta efter artikel om att längre vårdtis på IVA ger sämre utfall och
vad orsakerna är

@@

Sex, age and ASA preinjury were not significant in the unadjusted nor
the adjusted analyses, which implicates that the patients were not
treated differently based on age or gender. The results concerning the
ASA preinjury could be due to the OFI considering if each patient really
had an opportunity for improvement based on their individual cases,
meaning that if a patient has a higher ASA-score before the trauma, that
patient may not have the same reference point for OFI compared to a
patient with low ASA-score before the trauma.

#Lägga in artikel om hur IVA patienter behandlas baserat på kön eller
ålder + artikel om hur ASA-poängen eller pat allvarlighetsgrad påverkar
bedömningen av OFI

The variable on duty was not significant in either analyses indicating
that the time of arrival of the patients did not affect their medical
treatment. Trying to reduce number of OFI can therefore not be solved by
increasing medical resources outside working hours.

#Lägga in artikel som pekar på om duty påverkar handläggning eller inte

Something to consider is why ISS was not significant in the analyses.
Patients admitted to the ICU are usually in a bad shape with generally
high ISS. ISS is based on the square of the three highest AIS scores
allocated to the three most severly injured body parts. The downside of
ISS on its own is that it has higher predictive value when combined with
other parameters like vital signs, comorbidities and mechanism of
injury.

#Artikel om hur ISS bedöms kunna användas i samanhang av trauma

### Limitations

There were several limitations in our study: the selection for the
morbidity and mortality review, the definition of the RTS variable and
the study size.

The selection for review for the morbidity and mortality review relied
mostly on the local audit filters used at the Karolinska Institute
Hospital in Solna. This leads to a risk of misplacement where some
patients with OFI were not selected by for the review based on the audit
filters although there was an OFI later on in the hospitalization
process.

#Lägga in källa för audit filters - Kelvins artikel (mejlar Jocke)

The definition of the RTS-value is a combination of GCS, SBP and RR.
When the patients were intubated and had missing values for GCS or RR,
we replaced it with RTS with a value of 0 for the equation for RTS, as
GCS and respiratory rate are 0 when patients are intubated. This could
affect the final results, because then we would not get their initial
values for these variables, meaning some patients received a RTS score
more weighted towards SBP than they should have. However, the use of RTS
can also be seen as an advantage, being a combination of vital
parameters. This would mean that the RTS value reflects the state of the
patient better than these parameters separately.

The study was a single-centre study and illustrate the situation in the
Karolinska Intitute Hospital in Solna. More research is needed on the
area of OFI in patients admitted to the ICU. A suggestion would be to
conduct a multi-centre study on this group of patients to further
investigate areas for improvement and to further develop trauma quality
programs. Based on the results from this study, variables of interest
could be number of days in the ICU and investigation of GCS, RR and SBP
separately instead of them as a combined RTS-score. A study including
more patients could potentially show significance for the separate vital
parameters leading to a better prediction model for which patients that
should receive more focus when improving trauma care.

@@

The study is conducted on 143 outcomes of OFI. There were few outcomes
because we included only those who were admitted to the ICU, \> 14 years
old, not dead on arrival and had a registration for the presence or
absence of OFI. The external validity is therefore compromised due to
the limited size of the Swedish trauma database of patients admitted to
the ICU.

Based on the statistical significance for higher RTS being associated
with higher odds of OFI, there should be more focus on trauma patients
who are stable upon arrival who have a risk of becoming worse. Future
studies should investigate what factors to use to see the indication for
such a focus. Focusing on patients who are stable upon arrival overall
is very extensive. Therefore, identifying such factors would be
valuable.

@@

>>>>

## Bakgrund

Trauma är en ledande orsak till mortalitet och morbiditet i världen. En
hörnsten i utvecklingen av förbättringsprogram för trauma är
multidisciplinära mortalitets- och morbiditetskonferenser. Syftet med
dessa är att identifiera opportunities for improvement (OFI) som är en
viktig del i det riktade förbättringsarbetet av traumahandläggning.
Många patienter med svåra trauman blir inlagda på
intensivvårdsavdelningen (IVA). Trots detta är kännedomen om
förbättringsmöjligheter hos denna patientgrupp inte heltäckande i
nuläget. Syftet med denna studie är att karaktärisera OFI hos vuxna
traumapatienter som varit inlagda på IVA och bedöma hur patient- och
processfaktorer är associerade med OFI hos denna patientgrupp.

## Metod

Detta är en registerbaserad studie som inkluderar traumapatienter från
2014 till 2023 från Karolinska Universitetssjukhuset i Solna som har
varit inlagda på IVA och som har granskats avseende förekomsten av OFI
vid en mortalitets- och morbiditetskonferens. Bivariabel och
multivariabel logistisk regression användes för att bedöma sambandet
mellan följande patient- och processfaktorer samt OFI bland denna
patientgrupp: kön, ålder, om patienten var intuberad eller ej och hur
många dagar, Revised Trauma Score (RTS), Injury Severity Score (ISS),
jourtid vid ankomst till sjukhuset, tid från ankomst till sjukhuset till
första datortomografi (DT), dagar på IVA, ASA-värde före trauma och
överlevnad efter 30 dagar.

## Resultat

OFI identifierades hos 143 av 1449 patienter. Högre Revised Trauma Score
(RTS) var associerat med högre odds för OFI, ≤ 7 dagar på
intensivvårdsavdelningen (IVA) var associerat med lägre odds för OFI och
död inom 30 dagar efter sjukhusvistelse var associerat med högre odds
för OFI med statistisk signifikans i den ojusterade analysen. Högre
Revised Trauma Score (RTS) var associerat med högre odds för OFI, och ≤
7 dagar på IVA var associerat med lägre odds för OFI med statistisk
signifikans i den justerade analysen.

## Slutsats

De patient- och processfaktorer som var signifikant associerade med OFI
visade att högre RTS var associerat med högre odds för OFI och att ≤ 7
dagar på IVA var associerat med lägre odds för OFI.

@@

This was a registry-based study using all patients between 2014 and 2023
from the Karolinska University Hospital in Solna who had been admitted
to the ICU and who were reviewed regarding the presence of OFI by the
mortality and morbidity conference. Bivariable and multivariable
logistic regression was used to assess the following patient and process
factors and OFI among this group of patients: gender, age, if the
patient was intubated or not and how many days, Revised Trauma Score
(RTS), Injury Severity Score (ISS), arrival outside of working hours,
time from arrival at the hospital until first computed tomography (CT),
days in the ICU, ASA preinjury and survival after 30 days.

@@

APACHE = Acute Physiology and Chronic Health Evaluation

ASA score = American Society of Anesthesiologists score

@@

Albaaj et al. recently analyzed opportunities for improvement (OFI) in
relation to different patient and process factors based on a selection
of data from SweTrau including the patients admitted to the Karolinska
University Hospital in Solna following trauma. The study indicated that
patients with moderate to severe trauma are at the highest odds of OFI
[@4]. According to SweTrau 2022 41,1% of patients with New Injury
Severity Score (NISS) \> 15 who were at a university hospital were
admitted to the ICU. The proportion of severely injured admitted to the
ICU has decreased in recent years, however; patients with NISS \> 15
still constitute a large proportion of the trauma patients treated in
the ICU. A way to measure functional level after hospital discharge is
Glasgow Outcome Scale (GOS). According to GOS, among patients with NISS
\> 15 17% were deceased, 20% suffered severe disability, 49% suffered
moderate disability and 14% had good recovery in the university
hospitals of Sweden [@3].

@@

## Classifications systems for patient's general health

Another way to assess the state of a patient in addition to the injury
score systems are classification systems for patient's general health.
Two of these are ASA score and APACHE II.

American Society of Anesthesiologists (ASA) classification was
originally created to establish a scoring system for evaluation of a
patient's general health and comorbidities before an operative procedure
[@39]. However, recent studies have also showed ASA class to be a
significant predictor for length of stay, indicating that ASA score can
be used as predictor in other aspects of hospital care as well [@40].

Acute Physiology and Chronic Health Evaluation (APACHE) II is the most
commonly used severity-of-disease scoring system in ICUs all around the
world. The worst value for each physiological variabel is converted to
an integer score from 0 to 71 within the first 24 hours of admittance.
Higher score indicates more severe disease and higher risk for
mortality. The score has not been validated for use on patient's under
16 years of age. APACHE II has shown to be a good score for
discriminating between survivors and non-survivors [@41].

@@

For patients where hospital values for systolic blood pressure,
respiratory rate or GCS score were missing from the emergency
department, prehospital values were used instead. If the patient was
intubated we automatically assumed that GCS was equal to 3 and RR was
equal to 0. For the RTS-variable, each of the variables GCS, SBP and RR
were then converted to a RTS-code between 0 and 4 based on what the
values of GCS, SBP and RR were respectively. RTS was then calculated
using the formula for RTS based on the combination of the RTS-codes
assigned to each of the previously listed variables.

@@

The study outcome is on the presence of OFI in adult patients treated in
the ICU based on patient and process factors.

@@

The variables were factors chosen from the trauma registry, based on the
locally used audit filters, standard epidemiological factors and factors
registered in the Swedish Trauma registry. The categorical factors were
sex, age, if the patient was intubated or not and how many days, Revised
Trauma Score (RTS), Injury Severity Score (ISS), on call, time from
arrival at the hospital until first computed tomography (CT), days in
the ICU, ASA preinjury and survival after 30 days. On duty was defined
as before 8.00 a.m., after 5 p.m., or during a weekend - defined as
Saturday or Sunday.

@@

There were 14,022 patients in the trauma registry between January 1 2013
and February 1 2023. 2679 of these patients were admitted to the ICU. 3
patients were then excluded as they were younger than 15 years and 6
patients were excluded because they were dead on arrival. 928 patients
were excluded because there was no registration on whether there was an
OFI or not. 293 patients were excluded because they had missing values
in one or more variables. This leaves us with 143 outcomes for OFI, and
1,306 outcomes for "No OFI".

@@

Sex, Age, Intubation, ISS, time to first CT, on duty and ASA preinjury
were not significantly associated with OFI in the unadjusted nor the
adjusted analyses. Female sex, intubation \> 7 days, \> 7 days in the
ICU, ASA preinjury 1-2 and survival 30 days after hospitalisation could
not be analysed with logistic regression.

```{r, include=FALSE}
source("functions/table3.R")
```

```{r, echo=FALSE}
table3alive_merge
```

Table 3 shows the unadjusted and adjusted associations of OFI and the
selected patient and process factors among the patients who were alive
30 days after hospitalization. In the unadjusted analysis time to first
CT `r inline_text(table3aalive, variable = TimeFCT)` was significant.
Indicating that it has no correlation with increasing or decreasing odds
of OFI.

@@

The key results from this study are that we found that higher RTS was
associated with higher odds of OFI in both the adjusted and unadjusted
analyses, ≤ 7 days in the ICU was associated with decreased risk of OFI
in both the unadjusted and adjusted analyses, and death within 30 days
after hospitalization was associated with OFI in the unadjusted, but not
in the adjusted analysis. Age, gender, intubation, ISS, time to first
CT, on duty and ASA preinjury were not significantly associated with OFI
in neither the unadjusted or the adjusted analyses. In the subgroup
analysis of patients who were alive 30 days after hospitalization time
to first CT showed significance for having no relation to OFI.

We found that higher RTS was associated with higher odds of OFI. RTS is
a combined value of systolic blood pressure, Glasgow Coma Scale and
respiratory rate with more weight towards GCS. An explanation for the
increased odds of OFI in patients with increased RTS could be that the
patients with low RTS values - with low GCS, RR and SBP also are the
ones that are in a bad state upon arrival --- upon which emergency
procedures e.g. intubation have been carried out rapidly due to the
patients being visibly ill from the beginning. The patients who are
being missed and where certain procedures take longer time before being
performed are the patients who seem stable upon arrival, but who later
become worse --- the patients with comparatively higher RTS scores. In
Albaaj's study respiratory rate \>29, systolic blood pressure 50-75 mmHg
and 76-89 mmHg, and Glasgow Coma Scale 9-12 were significantly related
to increased odds of OFI in the unadjusted analysis, but not in the
adjusted analysis among patients from the part of the Swedish Trauma
Registry covering patients from the Karolinska University Hospital in
Solna [@12]. All of these variables equal medium to higher RTS values
--- capturing the group of patients who's vital parameters are
negatively affected by the trauma, but who are relatively stable in the
group of trauma patients. This creates a sense of false security,
leading to increased time until necessary emergency procedures are
performed.

The variable ≤ 7 days in the ICU was significantly associated with
decreasing odds of OFI in both the adjusted and unadjusted analyses.
There are two aspects to consider concerning this variable. One aspect
would be that less days spent in the ICU compared to more days spent in
the ICU indicates that patients have been in a better state and needed
less time of recovery before leaving the hospital. Furthermore, a longer
stay leads to more space for mistakes being made. The other aspect of
the variable would be that few days in the ICU could indicate the
patient passing away before any mistakes having been made due to the
patient being very ill from the beginning --- cases in which the deaths
were not preventable, but rather expected.

Previous studies have shown that prolonged ICU stay is a risk factor for
increased mortality and morbidity in trauma patients [@22, @26].
Something to consider is the reason for the lengthened stay in the ICU.
Studies have shown that patient characteristics on admission, like
comorbidities, age and severity of illness do not fully explain the
increased long-term mortality. With lengthened stay in the ICU the
reason for stay in the ICU become increasingly less associated with
patient's original reason for admittance and increasingly associated
with persistent chronic illness with associated organ dysfunction [@23,
@24]. This being the case, this might be an explanation for the lack of
sigificance for ASA preinjury and ISS in the analysis. As stated above
severity of illness and comorbidities do not fully explain increased
long-term mortality. OFI may therefore not correlate fully with ASA
preinjury and ISS respectively which are factors describing
comorbidities and the severity of illness.

However, a retrospective cohort has shown that lower admission illness
severity and younger age were associated with earlier discharge and
therefore less associated with persistent critical illness, which would
then indicate a stronger association between ISS and ASA preinjury to
length of stay in the ICU as the patients who are in a better state
leave the hospital earlier and there is less odds for OFI. On the other
hand, the cohort also showed that lower comorbidity was associated with
persistent critical illness rather than early discharge from the ICU ---
supporting the opposite side of the argument [@25].

Several studies have also studied the association between different
factors that affect the length of stay in the ICU. Associations between
hypoglycaemia, use of any corticosteroids, use of any neuromuscular
agents, use of any benzodiazepines for more than 1 day, the need for
mechanical ventilation for \> 2 days, initiation of new dialysis and the
occurrence of new infection has been found to have a correlation to
increased total and 5-year mortality [@23]. An association between
bleeding in patients and increased length of stay in the ICU has also
been found [@27].

@@

Sex and age were not significant in the unadjusted nor the adjusted
analyses, which implicates that the patients were not treated
differently based on age or gender. There are many studies on gender in
association to health care. It has previously been shown that there is
no difference in willingness to admit to the ICU between cases
describing a man or woman in the physician's responses [@29]. However,
studies also show that, among severely injured patients, female gender
is associated with lower in-hospital mortality among those aged 16-44
years. Female gender is also associated with decreased likelihood for
ICU admission [@30]. Male gender contributes more to poor outcomes in
postoperative ICU patients, and two-thirds of ICU resources are consumed
by male patients [@31]. However, in our study we could not see any
correlation between OFI and one specific gender in the occurrence of
OFI. Although we can also see that the majority of the patients included
in this study were also males. There is possibly a reason for higher
admittance of males to the ICU and association with poorer outcomes, but
it has nothing to do with odds of OFI.

ASA preinjury was not significant in the unadjusted nor the adjusted
analyses. Preinjury ASA score has been shown to be a reliable score for
predicting mortality after trauma, predicting readmission after
traumatic injury and for classifying comorbidity in trauma patients
[@32, @33, @34]. ASA score was originally developed to assess a
patient's health before surgical procedures and therefore provided
information on how much risk for postoperative complications there was.
Studies have investigated if ASA score also is a good predictor in other
aspects of medical care and as discussed above it is a predictor for
mortality after trauma and for readmission. However, this does not
automatically mean that there is an correlation between ASA score and
OFI. Patient's with higher ASA scores might receive more attention when
admitted to the ICU due to their comorbidities, but as the variable used
in this study is ASA preinjury --- a measurement of the patient's health
before the trauma --- this might not contribute to a fair picture of the
current state of the patient, which injury scores like RTS might.

The variable on duty was not significant in either analyses indicating
that the time of arrival of the patients did not affect their medical
treatment. Trying to reduce number of OFI can therefore not be solved by
increasing medical resources outside working hours. On-call sleeping has
been studied and it has been stated that sleeping between calls is
beneficial for waking performance. However, there might be a temporary
increase in performance impairment that should be taken into
consideration [@35].

Something to consider is why ISS was not significant in the analyses.
Patients admitted to the ICU are usually in a bad shape with generally
high ISS. ISS is based on the square of the three highest AIS scores
allocated to the three most severly injured body parts. The downside of
ISS on its own is that it has higher predictive value when combined with
other parameters like vital signs, comorbidities and mechanism of
injury. ISS is an anatomical model, compared to RTS which is a
physiological model. Among trauma patients admitted to the ICU the
physiological models perform better than the anatomical models. However,
patients with low scores in the physiological model also require an
anatomic analysis of the injuries to determine their severity [@36].
NISS has been found to be better than ISS in predicting the ICU
admission of trauma patients, as well as for predicting prolonged ICU
length of stay [@37]. This could be an explanation to why ISS is not
significant in the results of this analysis, as RTS as well as NISS are
potentially better models for patients admitted to the ICU.

### Strengths and limitations

One of the strengths of this study is that it is a registry-based study
which means that the quality of the data is reliable. The collection of
the included variables were pre-defined which means the reporting of the
included variables was standardized. Furthermore, the specific inclusion
criteria of the study: including patients 15 years and older, who were
not dead on arrival and who were admitted to the ICU, resulted in a
specific and well-defined cohort. This contributes to the reliability
and reproducibility of the study.

There were several limitations to our study: the limited study size, the
selection of patients for the morbidity and mortality review and the
definition and use of the physiological variables of the study.

Firstly, the study size was limited. This was a single-centre study and
illustrate the situation at the Karolinska Intitute Hospital in Solna.
Due to the small cohort of this study, it limits the applicability of
this study to other level 1 trauma centers and ICUs. The small cohort
might also affect the accuracy of the results.

Secondly, the selection of patients for the morbidity and mortality
review relied mostly on the local audit filters used at the Karolinska
Institute Hospital in Solna. Using audit filters to select patients for
the conference is associated with high false positive rates, ranging
from 24% to 80% [@38]. Even though the use of audit filters is
associated with high false positive rates, there is still a risk of
misplacement where some patients with OFI were not selected by for the
review based on the audit filters although there was an OFI that
occurred later on during the hospital stay.

The definition of the RTS-value is a combination of GCS, SBP and RR. The
variables used were the values from the emergency department. When these
were missing, the prehospital values were used. If these were also
missing, but there was documentation indicating that the patients had
been intubated, we replaced GCS and RR with a value of 0. This is
motivated for due to patients not breathing or being conscious when
intubated. The acquired values were then combined to calculate a value
for RTS. If the patient had missing values for either one of GCS, SBP or
RR they were excluded from the study. The handling of these variables
and their combination to RTS could affect the result as the values used
are from either prehospital values, emergency department values or from
identification of intubation. This could potentially affect the accuracy
of the data and decrease the possibility for reproducibility. However,
one could also argue that as the RTS value is a combined score of 3
physiological variables the combination of them decreases the overall
inaccuracy and produces a satisfactory measurement of the physiological
state of the patient.

@@

The study is conducted on 143 outcomes of OFI. There were few outcomes
because we included only those who were admitted to the ICU, \> 14 years
old, not dead on arrival and had a registration for the presence or
absence of OFI. The external validity is therefore compromised due to
the limited size of the study. The internal validity is however high
thanks to the well-defined cohort specifying trauma patients admitted to
the ICU at the Karolinska University Hospital in Solna. The results from
this study is therefore highly relevant for understanding where further
efforts for trauma improvement should be directed among ICU trauma
patients at this hospital.

### Implications

The potential clinical implication for this study is to further
contribute to the development of improved trauma care. Identifying the
group of patients admitted to the ICU is another step in the direction
of identifying the patient's who would benefit the most from further
improvements in trauma care. Trauma has a great global impact and this
study amongst others on the subject all contribute to the more specified
efforts in trauma quality improvement systems. Additionally, this study
has offered further insights to the patient profile of the trauma center
at Karolinska University Hospital in Solna admitted to the ICU.

### Future studies

More research is needed on the area of OFI in patients admitted to the
ICU. The most relevant limitation of this study was the limited study
size. A suggestion would be to conduct a multi-centre study, covering
larger databases or combining several databases, on this group of
patients to further investigate areas for improvement and to further
contribute to the improvement of trauma quality programs.

Based on the results from this study, variables of interest for future
studies could be number of days in the ICU and investigation of GCS, RR
and SBP separately instead of them as a combined RTS-score. It would be
interesting to see if one of the vital parameters is better for
predicting OFI than the others in a larger cohort.

Based on the statistical significance for higher RTS being associated
with higher odds of OFI, efforts should be more focused towards trauma
patients who are relatively stable upon arrival, but who have a risk of
getting worse. Future studies should investigate what factors that could
potentially indicate eventual worsening of the state of the patient.
Focusing on patients who are stable upon arrival overall is very
extensive, which further emphasizes the importance of identifying such
variables.

ASA preinjury did not show any correlation with OFI in this study.
However, it would be interesting to further investigate the correlation
between ASA preinjury or another scoring system of the patient's general
health before admittance to the hospital, and a combined physiological
and anatomical injury scoring system to see if they predict OFI better
combined than separately.

@@

# Contributions

This study has been conducted by Elin Sun Cao with continuous guidance
and feedback provided by Martin Gerdin Wärnberg, Jonatan Attergrim and
Kelvin Szolnoky.

# Acknowledgements

I want to extend my greatest gratitude to my supervisor Martin Gerdin
Wärnberg and co-supervisors Jonatan Attergrim and Kelvin Szolnoky who
have provided me with guidance and support throughout this semester.
Their extensive knowledge, deep insight into the subject and
availability and support has greatly contributed to the development of
this project. Additionally, I want to express my gratitude to my friends
and family who have been there contributing with encouragement and
positive energy.
